---
title: "Crime in Los Angeles"
output: 
  html_document: 
    fig_height: 4
    highlight: pygments
    theme: spacelab
---

## Setup

### Load packages
```{r load-packages, echo = FALSE, message = FALSE}
library(ggplot2)
library(dplyr)
library(statsr)
library(tidyr)
library(rmarkdown)
library(readr)
library(lubridate)
library(RColorBrewer)
library(ggrepel)
library(data.table)
```

### Load data
```{r load-data, echo = FALSE, message = FALSE}
# Set the Working Directory
setwd("Datasets")

# Import the full dataset as a dataframe
los_angeles_crime_data <- read_csv("Crime_Data_from_2020_to_Present.csv")

# Identify the character columns and convert them to factors if applicable.
char_cols <- sapply(los_angeles_crime_data, is.character)
los_angeles_crime_data[char_cols] <- lapply(los_angeles_crime_data[char_cols], as.factor)
```

* * *

## Part 1: Data 
##### Plan Phase of the PACE Framework

### How are the observations in the sample collected?
The data is obtained from the Los Angeles Police Department and provides valuable information about crime patterns and trends within the city. 
It is not possible to establish causal relationships as these are just observations and no controlled experiment was conducted.

Original data: https://data.lacity.org/Public-Safety/Crime-Data-from-2020-to-Present/2nrs-mtv8

* * *

## Part 2: Research question 
##### Plan Phase of the PACE Framework

### Project Scope (Specific if variables of interest are known, broad otherwise)
The purpose of this analysis is to extract insights from crime data in Los Angeles and explore the following questions:
- What is the actual average age of crime victims in Los Angeles, and how does it compare to the sample mean?
- Which areas of Los Angeles are considered the safest and most unsafe?
- Does a person's descent have any impact on their likelihood of being a crime victim?

* * *

## Part 3: Exploratory data analysis 
##### Analyze Phase of the PACE Framework

### Investigate the Dataset

#### Potential Areas to Investigate (if areas of investigation aren't already known)
- What is the true average age of crime victims in Los Angeles?
- What area of Los Angeles has the highest crime rate and at what time of day? (What area is most dangerous and least dangerous areas?)
- What role does sex play in crime victimization? (Are males or females more prone to crime victimization?)
- Does Ethnicity (Victim_Descent) play a role?

```{r eda}
# Initial investigation to identify trends and variables of interest
summary(los_angeles_crime_data)
head(los_angeles_crime_data)
tail(los_angeles_crime_data)

summary(los_angeles_crime_data$'Vict Descent')
summary(los_angeles_crime_data$'Vict Age')
summary(los_angeles_crime_data$'Vict Sex')
summary(los_angeles_crime_data$'AREA NAME')
summary(los_angeles_crime_data$'TIME OCC')

## Plot variables of interest

### Vict Descent
barplot(table(los_angeles_crime_data$'Vict Descent')) 
boxplot(table(los_angeles_crime_data$'Vict Descent'))
dotchart(table(los_angeles_crime_data$'Vict Descent'))
hist(table(los_angeles_crime_data$'Vict Descent'))

### Vict Age
hist(table(los_angeles_crime_data$'Vict Age'))
boxplot(los_angeles_crime_data$'Vict Age')
scatter.smooth(los_angeles_crime_data$'Vict Age')
qqnorm(los_angeles_crime_data$'Vict Age')
qqline(los_angeles_crime_data$'Vict Age')

### Vict Sex
barplot(table(los_angeles_crime_data$'Vict Sex')) 
boxplot(table(los_angeles_crime_data$'Vict Sex'))
hist(table(los_angeles_crime_data$'Vict Sex'))
dotchart(table(los_angeles_crime_data$'Vict Sex'))

### AREA NAME
barplot(table(los_angeles_crime_data$'AREA NAME'), las = 2)
boxplot(table(los_angeles_crime_data$'AREA NAME'))
hist(table(los_angeles_crime_data$'AREA NAME'))
dotchart(table(los_angeles_crime_data$'AREA NAME'))

### TIME OCC
hist(table(los_angeles_crime_data$'TIME OCC'))
boxplot(los_angeles_crime_data$'TIME OCC')
scatter.smooth(los_angeles_crime_data$'TIME OCC')
qqnorm(los_angeles_crime_data$'TIME OCC')
qqline(los_angeles_crime_data$'TIME OCC')

```

### Data Manipulation and Cleaning
```{r format_clean}
# Keep columns for analysis that were deemed relevant for the analysis
los_angeles_crime_data_trimmed <- los_angeles_crime_data %>%
  select(c("Vict Age","Vict Sex","Vict Descent","AREA NAME","TIME OCC"))

# Rename column headers for readability
colnames(los_angeles_crime_data_trimmed) <- c("Victim_Age", "Victim_Sex", "Victim_Descent", "Area_Name","Time_Occurred")

# Rename values in the Victim_Sex and Victim_Descent columns for readability.
los_angeles_crime_data_trimmed <- los_angeles_crime_data_trimmed %>%
  mutate(Victim_Sex = recode(Victim_Sex, "F" = "Female", "M" = "Male", "X" = "Unknown"),
         Victim_Descent = recode(Victim_Descent, "A" = "Other Asian", "B" = "Black", "C" = "Chinese", "D" = "Cambodian", "F" = "Filipino", "G" = "Guamanian", "H" = "Hispanic/Latin/Mexican", "I" = "American Indian/Alaskan Native",
  "J" = "Japanese", "K" = "Korean", "L" = "Laotian", "O" = "Other", "P" = "Pacific Islander", "S" = "Samoan", "U" = "Hawaiian", "V" = "Vietnamese", "W" = "White", "X" = "Unknown", "Z" = "Asian Indian"))

# Check for Missing Values
sum(is.na(los_angeles_crime_data_trimmed$Victim_Age))
sum(is.na(los_angeles_crime_data_trimmed$Victim_Sex))
sum(is.na(los_angeles_crime_data_trimmed$Victim_Descent))
sum(is.na(los_angeles_crime_data_trimmed$Area_Name))
sum(is.na(los_angeles_crime_data_trimmed$Time_Occurred))

# Omit rows with missing values
los_angeles_crime_data_trimmed <- na.omit(los_angeles_crime_data_trimmed)

# Filter out ages < 0 in the Victim Age column, only keep Male and Female in the Victim Sex column and remove Unknown from the Victim Descent Column.
los_angeles_crime_data_trimmed <- los_angeles_crime_data_trimmed %>%
  filter(Victim_Age > 0, Victim_Sex %in% c('Female','Male'), Victim_Descent != "Unknown")

if (FALSE) { # Skip this code
## Handle missing values through imputation
# Replace missing values in the Victim_Sex and Victim_Descent columns with the mode value.
mode <- names(which.max(table(los_angeles_crime_data_trimmed$Victim_Sex)))
los_angeles_crime_data_trimmed$Victim_Sex[is.na(los_angeles_crime_data_trimmed$Victim_Sex)] <- mode
mode <- names(which.max(table(los_angeles_crime_data_trimmed$Victim_Descent)))
los_angeles_crime_data_trimmed$Victim_Descent[is.na(los_angeles_crime_data_trimmed$Victim_Descent)] <- mode

# Replace 0's in the Victim_Age column with the mean age.
mean_vict_age <- mean(los_angeles_crime_data_trimmed$Victim_Age[los_angeles_crime_data_trimmed$Victim_Age != 0], na.rm = TRUE)
mean_vict_age <- round(mean_vict_age, digits = 0)
los_angeles_crime_data_trimmed$Victim_Age[los_angeles_crime_data_trimmed$Victim_Age == 0] <- mean_vict_age
}

# Convert the 24-hour values in the Time Occurred column to 12-hour.
los_angeles_crime_data_trimmed$Time_Occurred <- sprintf("%04d", as.integer(los_angeles_crime_data_trimmed$Time_Occurred))
los_angeles_crime_data_trimmed$Time_Occurred <- format(strptime(los_angeles_crime_data_trimmed$Time_Occurred, "%H%M"), "%I:%M %p")
los_angeles_crime_data_trimmed$Time_Occurred <- as.POSIXct(los_angeles_crime_data_trimmed$Time_Occurred, format = "%I:%M %p")

# View the data after cleaning
summary(los_angeles_crime_data_trimmed)
colnames(los_angeles_crime_data_trimmed)
los_angeles_crime_data_trimmed
```

### Analysis and Visualizations of Variables

#### When do most crimes occur?
According to this plot, there is a significant increase in crime at noon, while the 
occurrence of crimes is much lower at 4:00 AM. Between 4:00 PM and 7:00 PM, the 
rate of crime remains relatively constant.
```{r var_vis_one}
los_angeles_crime_data_trimmed %>%
  mutate(hourly_summary = format(Time_Occurred, format = "%H:00:00")) %>%
  group_by(hourly_summary) %>%
  summarize(count = n())

ggplot(los_angeles_crime_data_trimmed, aes(x = Time_Occurred)) +
  geom_line(stat = "bin", binwidth = 3600, color = "steelblue") +
  scale_x_datetime(date_labels = "%I:%M %p", date_breaks = "3 hours") +
  labs(x = "Time Occurred", y = "Count", title = "Crime in Los Angeles by Time of Day") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5),
        axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        axis.line = element_line(size = 0.5, color = "black"))

ggsave("Crime in Los Angeles by Time of Day.jpeg", width=6, height=8)
```

#### Geographical Region (Area_Name) vs Time of Day (Time_Occurred) and Sex (Victim_Sex)
Based on the data depicted in this plot, it can be inferred that the 77th Street and 
Central locations have a higher incidence of crime compared to other areas, which could 
make them less safe. It is worth noting that the occurrence of these crimes appears to 
be fairly consistent across all locations.
```{r var_vis_two}

ggplot(los_angeles_crime_data_trimmed, aes(x = Time_Occurred, fill = Victim_Sex)) +
  geom_histogram(binwidth = 60*60, color = "black", position = "stack") +
  facet_wrap(~Area_Name, ncol = 2) +
  scale_x_datetime(date_labels = "%I:%M %p", date_breaks = "3 hours") +
  labs(x = "Time Occurred", y = "Count", title = "Crime Distribution by Time and Gender by Area") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5),
        axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        legend.position = "top",
        legend.title = element_blank()) +
  scale_fill_manual(values = c("#1f77b4", "#ff7f0e"), labels = c("Male", "Female"))

ggsave("Crime Distribution by Time and Gender by Area.jpeg", width=6, height=8)
```

#### Victimization by Sex (Victim_Sex) and Geographical Region (Area_Name)
Based on this chart, it appears that both males and females in every area of Los Angeles 
are at similar risk of becoming victims of crime. However, the data suggests that females 
may be slightly more vulnerable. Certain regions, such as 77th Street and Central, demonstrate 
a higher incidence of crime compared to others.
```{r var_vis_three}
ggplot(los_angeles_crime_data_trimmed, aes(x = Area_Name, fill = Victim_Sex)) +
  geom_bar(position = "stack", color = "white") +
  labs(x = "Area Name", y = "Count", title = "Crime Distribution by Area and Sex") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5),
        axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        legend.position = "top",
        legend.title = element_blank()) +
  scale_fill_manual(values = c("#1f77b4", "#ff7f0e"), labels = c("Male", "Female")) +
  guides(fill = guide_legend(reverse = TRUE))

ggsave("Crime Distribution by Area and Sex.jpeg", width=6, height=8)
```

### Conditions
#### Test Type

```{r check_conditions}


```

* * *

## Part 4: Modeling/Inference 
##### Construct Phase of the PACE Framework
H0 (nothing going on):

HA (something going on):

```{r modeling_inference}
# Create a contingency table of the data
contingency_table <- table(gun_crime_data_trimmed$intent, gun_crime_data_trimmed$education)

# Run a Chi-Square Independence Test on the table.
chi_square_result <- chisq.test(contingency_table)

# Extract the test statistics
chi_square <- chi_square_result$statistic
df <- chi_square_result$parameter

# Set the desired confidence level
confidence_level <- 0.95

# Calculate the critical value
critical_value <- qchisq(1 - (1 - confidence_level) / 2, df)

# Calculate the margin of error
margin_of_error <- sqrt(chi_square / sum(chi_square_result$observed)) * critical_value

# Calculate the lower and upper bounds of the confidence interval
lower_bound <- chi_square / (1 - margin_of_error)
upper_bound <- chi_square / (1 + margin_of_error)
```


* * *

## Part 5: Prediction/Conclusion


```{r predictions_conclusion}
# Print the confidence interval
cat("Confidence Interval:", lower_bound, "-", upper_bound, "\n")

# Print the Chi Square Test Result
print(chi_square_result)
```

### What Story does the Data Tell?
##### Execute Phase of the PACE Framework
